Co-methylation

### 2015-05-04 Rmarkdown/comethylation.Rmd
Add plots highlighting bizarre chr21 results for within-fragment co-methylation

Another alternative to $\widehat{\theta}_{U}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$. This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods. The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta}_{U}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.

@@

Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_merged_MH_by_seqnames} show the results using CpG pairs with $NIL = 0$ for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_seqnames.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. All samples were sequenced with paired-end reads.}
\label{fig:Ziller_merged_MH_by_seqnames}
\end{figure}

@@

The equivalent plots for the $NIL \geq 0$ pairs paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_merged_ac_MH_by_seqnames}. In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_; _ADS_ is a female cell line), the _H9_-based samples (_H9_ and _H9\_Laurent_), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_; _IMR90_ is a female cell line).

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames.pdf}
\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. All samples were sequenced with paired-end reads.}
\label{fig:EPISCOPE_ac_MH_by_seqnames}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. All samples were sequenced with paired-end reads.}
\label{fig:Ziller_merged_ac_MH_by_seqnames}
\end{figure}

@@

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ pairs. Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI). All samples were sequenced with paired-end reads.}
\label{fig:Ziller_merged_MH_by_CGI}
\end{figure}

To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of pairs with $NIL \geq 0$ (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_merged_ac_MH_by_CGI}). We now see a fairly clear separation of the genome-wide estimates of $\theta$, with CpG islands being more co-methylated than the rest of the genome. Moreover, the estimates of $\theta$ within CpG islands are approximately constant with respect to $IPD$. Also notable is that the pairs of CpGs outside of CpG islands have fairly similar estimates of $\theta$ in both the $NIL = 0$ and $NIL \geq 0$ analyses.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_CGI.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs. Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI). All samples were sequenced with paired-end reads.}
\label{fig:Ziller_merged_ac_MH_by_CGI}
\end{figure}

>>>>

Another alternative to $\widehat{\theta}_{U}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$. This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods. The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta}_{U}$, when the sample size, $n$, is small \citep[pp. 157-158]{Agresti:2007tf}.

@@

Figure \ref{fig:EPISCOPE_MH_by_seqnames}, \ref{fig:Lister_MH_by_seqnames}, \ref{fig:Seisenberger_MH_by_seqnames} and \ref{fig:Ziller_MH_by_seqnames} show the results using CpG pairs with $NIL = 0$ for the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets, respectively.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_seqnames.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. All samples were sequenced with paired-end reads.}
\label{fig:Ziller_MH_by_seqnames}
\end{figure}

@@

The equivalent plots for the $NIL \geq 0$ pairs paint a similar picture, as shown in Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames}, \ref{fig:Lister_ac_MH_by_seqnames}, \ref{fig:Seisenberger_ac_MH_by_seqnames} and \ref{fig:Ziller_ac_MH_by_seqnames}. In addition to the mitochondria, the estimates for X chromosome are frequently distinguishable from the rest of the chromosome-level estimates; e.g., the _ADS_-based samples (_ADS_, _ADS-adipose_ and _ADS-iPSC_; _ADS_ is a female cell line), the _H9_-based samples (_H9_ and _H9\_Laurent_), the _IMR90_-based samples (_IMR90\_r1_, _IMR90\_r2_, _IMR90_cell_line_, _IMR90-iPSC_; _IMR90_ is a female cell line).

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames.pdf}
\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. All samples were sequenced with paired-end reads. There is no Y chromosome or mtDNA data for \emph{E18BUF} due to a coding error that meant these chromosomes weren't processed by \texttt{methtuple} for \texttt{2ac} tuples. This omission does not affect the other results.}
\label{fig:EPISCOPE_ac_MH_by_seqnames}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. All samples were sequenced with paired-end reads.}
\label{fig:Ziller_ac_MH_by_seqnames}
\end{figure}

@@

In many of the $NIL \geq 0$ plots there are samples with autosomal $\widehat{\theta}_{MH}$ that are noticeably separated from the main cloud of autosomal estimates. For example, this can be readily observed in the _Colon_Primary_Normal_, _Colon_Primary_Tumour_ samples from the _Ziller_ dataset (Figure \ref{fig:Ziller_ac_MH_by_seqnames}), but it is also apparent in other samples from the _Ziller_ dataset and the _EPISCOPE_ dataset (Figure \ref{fig:EPISCOPE_ac_MH_by_seqnames}). These plots do not distinguish estimates from different autosomes. Further analysis, however, reveals that these outliers almost all come from a single chromosome, chromosome 21.

Figures \ref{fig:EPISCOPE_ac_MH_by_seqnames_chr21} (_EPISCOPE_), \ref{fig:Lister_ac_MH_by_seqnames_chr21} (_Lister_) and \ref{fig:Ziller_ac_MH_by_seqnames_chr21} (_Ziller_) show the exact same plots as before, but with the chromosome 21 data highlighted by a black line. The chromosome 21 data stand out remarkably from the rest of the autosomal data for all samples from the _EPISCOPE_ dataset and several samples from the _Ziller_ dataset. Just as remarkable, however, is that this phenomenon is not observed in the _Lister_ dataset nor for any of the frontal cortex samples form the _Ziller_ dataset[^mice_autosomes]. At this time I have no explanation for this result. I cannot rule out it being a technical artefact, although preliminary analyses suggest this is not the case (data not shown), and I do not have a hypothesis for a potential biological cause of this phenomenon.

[^mice_autosomes]: This phenomenon cannot occur in the _Seisenberger_ dataset since these samples are mice, and mice only have 19 autosomes.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/EPISCOPE_ac_MH_by_seqnames_chr21.pdf}
\caption[\emph{EPISCOPE}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{EPISCOPE} dataset using $NIL \geq 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. The chromosome 21 data are shown by the black line. All samples were sequenced with paired-end reads. There is no Y chromosome or mtDNA data for \emph{E18BUF} due to a coding error that meant these chromosomes weren't processed by \texttt{methtuple} for \texttt{2ac} tuples. This omission does not affect the other results.}
\label{fig:EPISCOPE_ac_MH_by_seqnames_chr21}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_ac_MH_by_seqnames_chr21.pdf}
\caption[\emph{Lister}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Lister} dataset using $NIL \geq 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. The chromosome 21 data are shown by the black line. Only the \emph{ADS}, \emph{ADS-adipose}, \emph{ADS-iPSC} and \emph{H9\_Laurent} samples were sequenced with paired-end reads.}
\label{fig:Lister_ac_MH_by_seqnames_chr21}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_seqnames_chr21.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs with chr21 highlighted]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs. The genome-level estimates are shown by the blue line while the chromosome-level estimates are plotted as points coloured by whether the chromosome is an autosome, X chromosome, Y chromosome or mtDNA. The chromosome 21 data are shown by the black line. All samples were sequenced with paired-end reads.}
\label{fig:Ziller_ac_MH_by_seqnames_chr21}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_MH_by_CGI.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL = 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL = 0$ pairs. Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI). All samples were sequenced with paired-end reads.}
\label{fig:Ziller_MH_by_CGI}
\end{figure}

To look for evidence of higher-order within-fragment co-methylation we turn to the analysis of pairs with $NIL \geq 0$ (Figures \ref{fig:EPISCOPE_ac_MH_by_CGI}, \ref{fig:Lister_ac_MH_by_CGI}, \ref{fig:Seisenberger_ac_MH_by_CGI} and \ref{fig:Ziller_ac_MH_by_CGI}). We now see a fairly clear separation of the genome-wide estimates of $\theta$, with CpG islands being more co-methylated than the rest of the genome. Moreover, the estimates of $\theta$ within CpG islands are approximately constant with respect to $IPD$. Also notable is that the pairs of CpGs outside of CpG islands have fairly similar estimates of $\theta$ in both the $NIL = 0$ and $NIL \geq 0$ analyses.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Ziller_merged_ac_MH_by_CGI.pdf}
\caption[\emph{Ziller}: $\widehat{\theta}_{MH}$ for $NIL \geq 0$ CpG pairs stratified by CpG islands]{The Mantel-Haenszel estimate of the log odds ratio, $\widehat{\theta}_{MH}$, is plotted as a function of $IPD$ for the \emph{Ziller} dataset using $NIL \geq 0$ pairs. Only the genome-level estimates are shown, stratified by whether the pair is inside a CpG island (CGI). All samples were sequenced with paired-end reads.}
\label{fig:Ziller_ac_MH_by_CGI}
\end{figure}
